---
title: "GSE47552 Microarray Analysis"
author: "Popoola Abdulgafar"
date: "`r Sys.Date()`"
output: html_document
---

# GSE47552 Microarray Analysis

## 1. Introduction

This project analyzes the GSE47552 gene-expression dataset to identify genes that are differentially expressed between Multiple Myeloma (MM) plasma cells and Normal Plasma Cells (NPC).

The analysis includes differential expression analysis, visualization of expression patterns, principal component analysis (PCA), heatmap analysis, and functional enrichment using Gene Ontology (GO) and KEGG pathways.

## 2. Dataset

The dataset was obtained from the Gene Expression Omnibus (GEO) using the GEOquery package.

The analysis focuses on the comparison between Multiple Myeloma (MM) plasma cells and Normal Plasma Cells (NPC).

## 3. Load packages and data

```{r setup}
library(GEOquery)
library(limma)
library(ggplot2)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)

gse <- getGEO("GSE47552")

expr <- exprs(gse[[1]])
pheno <- pData(gse[[1]])

dim(expr)
```
## 4. Sample metadata

The sample metadata were examined to identify the biological groups represented in the dataset.

```{r metadata}
metadata <- data.frame(
  Sample = rownames(pheno),
  Group = pheno$`cell type:ch1`
)

table(metadata$Group)
head(metadata)
```
## 5. Differential expression: MM versus NPC

For the differential-expression analysis, Multiple Myeloma (MM) plasma cells were compared with Normal Plasma Cells (NPC). The analysis was performed using the `limma` package, which is appropriate for microarray expression data.

```{r differential-expression}
keep <- metadata$Group %in% c(
  "Multiple Myeloma (MM) plasma cells",
  "Normal Plama Cells (NPC)"
)

expr_sub <- expr[, keep]
meta_sub <- metadata[keep, ]

meta_sub$Group <- factor(
  meta_sub$Group,
  levels = c(
    "Normal Plama Cells (NPC)",
    "Multiple Myeloma (MM) plasma cells"
  )
)

design <- model.matrix(~0 + meta_sub$Group)
colnames(design) <- c("NPC", "MM")

fit <- lmFit(expr_sub, design)

contrast.matrix <- makeContrasts(
  MMvsNPC = MM - NPC,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

results <- topTable(
  fit2,
  coef = "MMvsNPC",
  number = Inf,
  adjust.method = "BH"
)

head(results)
```
## 6. Volcano plot

A volcano plot was used to visualize the relationship between statistical significance and magnitude of differential expression. The x-axis represents log fold change, while the y-axis represents the negative log10-transformed P-value.

```{r volcano}
results$Significant <- ifelse(
  results$adj.P.Val < 0.05 & abs(results$logFC) > 1,
  "Significant",
  "Not significant"
)

ggplot(results, aes(x = logFC, y = -log10(P.Value))) +
  geom_point(aes(color = Significant), alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Volcano Plot: MM vs NPC",
    x = "log2 Fold Change",
    y = "-log10(P-value)"
  )
```
## 7. Principal Component Analysis (PCA)

Principal component analysis was used to visualize the overall structure of the gene-expression data and assess whether MM and NPC samples show separation based on their global expression profiles.

```{r PCA}
pca <- prcomp(t(expr_sub), scale. = TRUE)

pca_data <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  Group = meta_sub$Group
)

ggplot(pca_data, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "PCA of MM and NPC Samples",
    x = "PC1",
    y = "PC2"
  )
```
## 8. Heatmap of differentially expressed genes

A heatmap was used to visualize the expression patterns of selected differentially expressed genes across the MM and NPC samples.

```{r heatmap}
sig_genes <- results[
  results$adj.P.Val < 0.05 & abs(results$logFC) > 1,
]

top_genes <- rownames(sig_genes)[1:min(50, nrow(sig_genes))]

heatmap_data <- expr_sub[top_genes, ]

pheatmap(
  heatmap_data,
  scale = "row",
  show_rownames = FALSE,
  annotation_col = data.frame(
    Group = meta_sub$Group
  )
)
```
## 9. Gene Ontology (GO) enrichment

Gene Ontology enrichment was performed to identify biological processes and molecular functions associated with the differentially expressed genes.

```{r GO-enrichment}
# Extract significant probe IDs
sig_probe <- rownames(sig_genes)

# Map probes to gene symbols using the GPL annotation
gene_assignment <- gpl_table$gene_assignment[
  match(sig_probe, gpl_table$ID)
]

# Extract gene symbols
gene_symbols <- sapply(
  strsplit(gene_assignment, "///"),
  function(x) {
    if (length(x) == 0 || x[1] == "---") return(NA)

    parts <- trimws(strsplit(x[1], "//", fixed = TRUE)[[1]])

    if (length(parts) >= 2) parts[2] else NA
  }
)

gene_symbols <- unique(na.omit(gene_symbols))

# Convert gene symbols to Entrez IDs
gene_ids <- bitr(
  gene_symbols,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

# GO enrichment
ego <- enrichGO(
  gene = gene_ids$ENTREZID,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  pAdjustMethod = "BH",
  readable = TRUE
)

dotplot(ego, showCategory = 20)
```
## 10. KEGG pathway enrichment

KEGG enrichment was used to identify biological pathways represented among the differentially expressed genes.

```{r KEGG-enrichment}
ekegg <- enrichKEGG(
  gene = gene_ids$ENTREZID,
  organism = "hsa",
  pAdjustMethod = "BH"
)

dotplot(ekegg, showCategory = 20)
```
